LabMate: supporting types for MATLAB

### 2024-04-08 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] added typing rules figure

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
\usepackage{proof}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{subcaption}
%\usepackage{multirow}
\usepackage{xkeyval}
\usepackage[font=small,justification=centering]{caption}
\captionsetup{labelsep = period}
\usepackage{floatrow}

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Summarised as an inference rule, matrix types are thus formed as follows:
%
\[
  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
  {
    \istype{R}
    &\quad
    \istype{C}
    &\quad
    x : R, y : C \vdash \istype{E(x,y)}
    &\quad
    rs : \List{R}
    &\quad
    cs : \List{C}
  }
\]

@@

Two matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
numeric (for example integers, or floating point numbers), and if $A$
and $B$ have the same matrix type:
%
\[
  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
  {
    x : R, y : C \vdash \isadd{E(x, y)}
    &\quad
    A : \Matrix{R}{C}{E}{rs}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{rs}{cs}
  }
\]
%
Matrix multiplication is trickier, because it does not demand that $A$
and $B$ have the same type --- rather, it demands that $A$ and $B$
have \remph{compatible} types:
%
\[
  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
  {
    \ismult{E}{E'}{E''}
    &\quad
    A : \Matrix{R}{C}{E}{rs}{ms}
    &\quad
    B : \Matrix{R}{C}{E'}{ms}{cs}
  }
\]
The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
multiplication operation with codomain $E''(x, z)$.
%
This is certainly the case if all entry types involved are simply
integers or floating point numbers, but this constraint can also
ensure that the physical units involved line up.

@@

Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
\begin{verbatim}
%> A :: [ i <- 12,  j <- 5 ] (if j == 4 then Q T else Q 1)
%> B :: [ i <- 5 ] (if i == 4 then Q (ML^2T^(-4)I^(-1)) else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities. Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively. LabMate reports that the type of \texttt{C} should be
\begin{verbatim}
%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$. Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
\begin{verbatim}
%> B :: [ Q TODO; Q TODO; ?aTy; Q TODO; Q TODO; Q TODO ]
\end{verbatim}
We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}. This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.

>>>>

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
\usepackage{proof}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{subcaption}
%\usepackage{multirow}
\usepackage{xkeyval}
\usepackage[font=small,justification=centering]{caption}
\captionsetup{labelsep = period}
\usepackage{subcaption}
\usepackage{floatrow}

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        rs : \List{R}
        &\quad
        cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{rs}{cs}
        &\quad
        B : \Matrix{R}{C}{E}{rs}{cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{rs}{ms}
        &\quad
        B : \Matrix{R}{C}{E'}{ms}{cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

@@

The conditions for adding two matrices are given in Figure~\ref{fig:matrix_add} --- the matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
numeric (for example integers, or floating point numbers), and if $A$
and $B$ have the same matrix type.
%
%\[
%  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
%  {
%    x : R, y : C \vdash \isadd{E(x, y)}
%    &\quad
%    A : \Matrix{R}{C}{E}{rs}{cs}
%    &\quad
%    B : \Matrix{R}{C}{E}{rs}{cs}
%  }
%\]
%

Matrix multiplication is trickier, because it does not demand that $A$
and $B$ have the same type --- rather, it demands that $A$ and $B$
have \remph{compatible} types, see Figure~\ref{fig:matrix_mul}.
%
%\[
%  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
%  {
%    \ismult{E}{E'}{E''}
%    &\quad
%    A : \Matrix{R}{C}{E}{rs}{ms}
%    &\quad
%    B : \Matrix{R}{C}{E'}{ms}{cs}
%  }
%\]
The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
multiplication operation with codomain $E''(x, z)$.
%
This is certainly the case if all entry types involved are simply
integers or floating point numbers, but this constraint can also
ensure that the physical units involved line up.

@@

Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
\begin{verbatim}
% > A :: [ i <- 12,  j <- 5 ]
      (if j == 4 then Q T else Q 1)
% > B :: [ i <- 5 ]
      (if i == 4
         then Q (ML^2T^(-4)I^(-1))
         else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities. Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively. LabMate reports that the type of \texttt{C} should be
\begin{verbatim}
%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$. Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
\begin{verbatim}
% > B :: [ Q TODO; Q TODO; ?aTy;
           Q TODO; Q TODO ]
\end{verbatim}
We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}. This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.
